---
title: "Simulations Pre-Class Project"
date: "Due March 13, 2017 at 5:00pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:


With this project we will simulate a famoues probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
    **ANSWER** 8.74%
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    **ANSWER** 42.3%
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
    **ANSWER** On average our player goes "bust" on the 505th hand (conditional on busting within first 5,000 hands)
    d. the mean and variance of your bankroll after 100 hands (including busts).
    **NOTE** This question requires _so_ much data to be retained (if you want to output the results) that I feel my approach may have been incorrect but here's my answer.
    **ANSWER** The mean bankroll after 100 hands (making busts = 0), is 211 hands while the sd of the bankrolls is 546!
    e. the mean and variance of your bankroll after 500 hands (including busts).
    **ANSWER** The mean is 90.6 and sd is 383. 
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

``` {r echo = TRUE} 
# First attempt to simulate the Gambler's Ruin 
# Seems like we want to use a binomial distribution, since we have "n" independent trials where the result is either success (you won) or failure (you lost), according to a constant probability of success. This is NOT yet a Markov Chain since you do not change your strategy depending on wins / losses

set.seed(1) # for reproducibility

initialBankroll <- 1000 # start with $1,000 but will increment this
bet <- 100 # always bet $100 
pSuccess <- .5 # I assume a "fair" games means 50% chance to win 
numIteractions <- 5000 # could change for other simulations

gamblerSimulation <- function(initialBankroll = 1000, bet = 100, pSuccess = .5, numIteractions = 5000) {

  bankrollHistory <- rep(NA, numIteractions) # create a bankroll History
  bankrollHistory[1] <- initialBankroll # set beginning bankroll 
  
  for (i in 1:numIteractions) { # i is the "hand" the game is on (first, second, etc.)
    
    # Check whether our player can even play the game! 
    if (bankrollHistory[i] < 100) { # Our player doesn't have enough money to play
      output <- list("Bust", i, bankrollHistory)
      return(output)
    } else { # our player does have enough money to play 
      # Have our player play the hand with a 50% probability of winning 
      trialSuccess <- rbinom(1, 1, p = pSuccess) # I want to sample 1 pull 1 time, with a probability of 50% of a 1 or 0
      
      if (trialSuccess == 1) { # won!
        bankrollHistory[i+1] <- bankrollHistory[i] + bet 
      } else { # lost
        bankrollHistory[i+1] <- bankrollHistory[i] - bet
      }
    }
  }
  output <- list("Not Bust", i, bankrollHistory)
  return(output)
}

# Simulate 5,000 versions of the game (with max 5,000 hands) 
simulations <- replicate(numIteractions, gamblerSimulation())

# To answer the first three questions easily, I'll extract only the "Bust" and "Hand" variables from the list simulations (leaving the Bankroll variable for later)
simulationsReshaped <- data.frame(matrix(unlist(simulations[1:2,]), nrow = numIteractions, byrow = T))

colnames(simulationsReshaped) <- c("Busted", "Hand", "Bankroll")
simulationsReshaped$Hand <- as.numeric(simulationsReshaped$Hand)
simulationsReshaped$Bankroll <- as.numeric(simulationsReshaped$Bankroll)

## Answering subquestions 
library(dplyr)
answerA <- simulationsReshaped %>%
  filter((Busted == "Bust" & Hand <= 100)) %>%
  summarise(finalAnswer = n() / numIteractions * 100)

answerB <- simulationsReshaped %>%
  filter((Busted == "Bust" & Hand <= 500)) %>%
  summarise(finalAnswer = n() / numIteractions * 100)

answerC <- simulationsReshaped %>%
  filter(Busted == "Bust") %>% # only include simulations where the player Busted
  summarise(finalAnswer = mean(Hand))

# Answering questions d and e requires a little different data (i.e., the entire bankroll history for each hand) which is ridiculously large 
# I would like to learn the most efficiency way to do this in the future please!!!
answerD <- unlist(simulations[3,][100]) # should pull out the bankroll (3) and [100] specifies only the 100th element of each bankroll
answerD[is.na(answerD)] <- 0 # convert NA (which means they busted) to 0 to calculate mean & var

mean(answerD)
sd(answerD)

answerE <- unlist(simulations[3,][500]) # should pull out the bankroll (3) and [100] specifies only the 100th element of each bankroll
answerE[is.na(answerE)] <- 0 # convert NA (which means they busted) to 0 to calculate mean & var

mean(answerE)
sd(answerE)
```

2. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
