---
title: "pre-class-03"
author: "Joey Heffner"
date: "9/21/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
```

## Setup

```{r}
#install.packages("gapminder")
library(dplyr)
library(gapminder)
gapminder # little risky to open a dataset without head() if you don't know it's a tibble!
```

## Pre-class problems

1. How many unique countries are represented per continent? 

**ANSWER.** 52 for Africa, 25 for Americas, 33 for Asia, 30 for Europe, 2 for Oceania

```{r}
dfUniqueCountries <- gapminder %>%
  group_by(continent) %>% 
  summarise(uniqueCountries = n_distinct(country))
dfUniqueCountries
```

2. Which European nation had the lowest GDP per capita in 1997? 

**ANSWER.** Albania.

```{r}
gdp1997 <- gapminder %>%
  filter(continent == "Europe" & year == 1997) %>% 
  arrange(gdpPercap)
gdp1997
```

3. According to the data available, what was the average life expectancy across each continent in the 1980s? 

**ANSWER.** See avgLifeExpExp1980s tibble. 

```{r}
avgLifeExp1980s <- gapminder %>% 
  filter(year %in% c(1980:1989)) %>%
  group_by(continent) %>% 
  summarise(avgLifeExp = mean(lifeExp))
avgLifeExp1980s
```

4. What 5 countries have the highest total GDP over all years combined? 

**ANSWER.** Kuwait, Switzerland, Norway, United States, and Canada. 

```{r}
totalGDP <- gapminder %>%
  group_by(country) %>% 
  summarise(totalGDP = mean(gdpPercap)) %>%
  arrange(desc(totalGDP))
totalGDP
```

5. What countries and years had life expectancies of *at least* 80 years? *N. B. only output the columns of interest: country, life expectancy and year (in that order).*

**ANSWER.** See lifeExp80Yrs. Question is a little ambiguous so I coded it to include all possible combinations of country & year.

```{r}
lifeExp80Yrs <- gapminder %>%
  select(country, lifeExp, year) %>%
  filter(lifeExp >= 80)
lifeExp80Yrs
```

6. What 10 countries have the strongest correlation (in either direction) between life expectancy and per capita GDP?

**ANSWER.** France, Austria, Belgium, Norway, Oman, United Kingdom, Italy, Isreal, Denmark, and Australia. 

```{r}
lifeExpGDP <- gapminder %>%
  group_by(country) %>% 
  summarise(corLifeGDP = cor(lifeExp, gdpPercap)) %>%
  arrange(desc(abs(corLifeGDP)))
lifeExpGDP
```

7. Which combinations of continent (besides Asia) and year have the highest average population across all countries? *N. B. your output should include all results sorted by highest average population.* With what you already know, this one may stump you. See this Q&A for how to `r ungroup` before `r arrange`ing. This also behaves differently in more recent versions of dplyr.

**ANSWER.** Americas in 2007 had the highest average population (35,954,847)

```{r}
avgPop <- gapminder %>%
  filter(continent != "Asia") %>% # remove Asia from the tibble
  group_by(continent, year) %>% # group by every combination of continent and year
  summarise(avgPop = mean(pop)) %>% # average population (across all countries) for each combination of continent and year
  ungroup() %>% # is this necessary here? tibble ends up the same with or without this ungroup()
  arrange(desc(avgPop)) 
avgPop
```

8. Which three countries have had the most consistent population estimates (i.e. lowest standard deviation) across the years of available data?

**ANSWER.** Sao Tome and Principe, Iceland, and Montenegro

```{r}
populationSD <- gapminder %>%
  group_by(country) %>%
  summarise(popSD = sd(pop)) %>% # summarise standard deviation of population for all countries across all years
  arrange(popSD)
populationSD
```

9. Subset **gm** to only include observations from 1992 and store the results as gm1992. What kind of object is this?

**ANSWER.** It's still a tibble like `r class(gapminder)` shows as well. 

```{r}
gm1992 <- gapminder %>%
  filter(year == 1992)
class(gm1992)
```

10. Which observations indicate that the population of a country has *decreased* from the previous year **and** the life expectancy has *increased* from the previous year? See the vignette on window functions. 

**ANSWER.** The question is a little ambiguous again because there's actually multiple observations showing that the population of a country decreased (from the previous year) and the life expectancy increased (from the previous year). Based on my (hopefully right) answer there would be 36 observations (see final printed tibble) where the population decreased from the prior avaliable year but the life expectancy increased. 

```{r}
specificObs <- gapminder %>%
  group_by(country) %>% # grouping by country 
  mutate(popDelta = pop - lag(pop, order_by = year)) %>% # difference in pop by year
  mutate(lifeExpDelta = lifeExp - lag(lifeExp, order_by = year)) %>% # difference in life exp by year 
  arrange(country, year) # not strictly necessary but useful to visualize

specificObs %>% 
  filter(popDelta < 0 & lifeExpDelta >0) # filtering based on conditions specified

```
