---
title: "pre-class-05"
author: "Joey Heffner"
date: "October 9, 2018"
output: html_document
---

# pre-class
Make sure you commit this often with meaningful messages. 

**Manual Commit / Comment:** I completed this assignment after class so the same answer is in "in-class-04.Rmd."

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Standardizing a variable means subtracting the mean, and then dividing by the standard deviation. Let’s use a loop to standardize the numeric columns in the [Western Collaborative Group Study](https://clinicaltrials.gov/ct2/show/NCT00005174). This study began in 1960 with 3154 men ages 39-59, who were employed in one of 11 California based companies. They were followed until 1969 during this time, 257 of these men developed coronary heart disease (CHD). You can read this data in with the code below. You can access this dataset with the following code:

``` {r echo = TRUE}
suppressMessages(library(foreign))
wcgs <- read.dta("https://drive.google.com/uc?export=download&id=0B8CsRLdwqzbzYWxfN3ExQllBQkU")
```

The data has the following variables:

WCGS has the following variables:

-----------------------------------------------------------
Name    Description
------- -------------------------------------------
id      Subject identification number

age     Age in years

height  Height in inches

weight  Weight in lbs.

sbp     Systolic blood pressure in mm Hg

dbp     Diastolic blood pressure in mm Hg

chol    Fasting serum cholesterol in mm 

behpat  Behavior

  1       A1

  2       A2

  3       B3

  4       B4

ncigs   Cigarettes per day

dibpat  Behavior

1       type A

2       type B

chd69   Coronary heart disease

1       Yes

0       no

typechd Type of CHD

1       myocardial infarction or death

2       silent myocardial infarction

3       angina perctoris

time169 Time of CHD event or end of follow-up

arcus   Arcus senilis

0       absent

1       present

bmi     Body Mass Index
-----------------------------------------------------------


### Question 1: Standardize Function

A. Create a function called standardize.me() that takes a numeric vector as an argument, and returns the standardized version of the vector. 

``` {r echo = TRUE}

standardize.me <- function(x) {
 #Standardizing a variable means subtracting the mean, and then dividing by the standard deviation 
  
  mean.x <- mean(x, na.rm = TRUE) 
  sd.x <- sd(x, na.rm = TRUE)

  std.x <- NULL # initialize
  for (i in 1:length(x)) { # not sure if a for loop is necessary to do this but seems obvious start
    if (is.na(x[i])) { # had to add because one of the cols had an NA
      next
    } else {
    std.x[i] <- (x[i] - mean.x) / sd.x
    }
  }
  
  return(std.x)
}

# Test function
x <- rnorm(100, mean = 50, sd = 10) # generate normal distribution numbers 

x1 <- standardize.me(x) # testing my standardize function
x2 <- scale(x) # testing R's scale function
x3 <- mapply(x2, FUN=as.numeric) # convert to a vector to compare to x1

identical(x1, x3) # hurray!
```

B. Assign all the numeric columns of the original WCGS dataset to a new dataset called WCGS.new.

``` {r echo = TRUE}

# since wcgs is a dataframe and thus a list we can use list-apply functions to select numeric columns only:
subsetIndex <- unlist(lapply(wcgs, is.numeric)) # creates logicals specifying numeric columns
                                                # knowing what I know now (after Datacamp) I would use the map functions here instead of lapply in the future
wcgs.new <- wcgs[,subsetIndex] # selects correct columns based on logical index
```

C. Using a loop and your new function, standardize all the variables WCGS.new dataset.

``` {r echo = TRUE}

# IMPORTANT: Standardize function must be able to accomodate NAs in some way (I chose to ignore and not update NAs but you could interpolate or use some other method)

for (j in 1:ncol(wcgs.new)) { # loops through number of cols
  colVariable <- wcgs.new[,j] # pull vector from col
  wcgs.new[,j] <- standardize.me(colVariable) # standardize variables and put them back in dataframe
}

```

D. What should the mean and standard deviation of all your new standardized variables be? Test your prediction by running a loop

**ANSWER** Mean should be 0 and SD should be 1. sumData shows that!

``` {r echo = TRUE}

sumData <- matrix(ncol = ncol(wcgs.new), nrow = 2)

for (j in 1:ncol(wcgs.new)) {
  sumData[1,j] <- mean(wcgs.new[,j])
  sumData[2,j] <- sd(wcgs.new[,j])
}

sumData

```

### Question 2: Looping to Calculate

A. Using a loop, calculate the mean weight of the subjects separated by the type of CHD they have.

``` {r echo = TRUE}
# I assume we're suppose to use the original wcgs dataframe for this (not the standardized version)
# I had a hard time coming up with an efficient way to do this (without using dplyr) so it uses a lot of if statements!

uniqueTypch69 <- unique(wcgs["typchd69"]) # grabbing the unique labels

noCHDWeight <- NULL
miOrSDWeight <- NULL
silentMIWeight <- NULL
anginaWeight <- NULL

for (j in 1:nrow(wcgs)) { # go through each row
  if (wcgs[j,"typchd69"] == uniqueTypch69[1,1]) { # if matches first label
    noCHDWeight <- append(noCHDWeight, wcgs[j, "weight"])
  } else if (wcgs[j,"typchd69"] == uniqueTypch69[2,1]) { # matches second label
    miOrSDWeight <- append(miOrSDWeight, wcgs[j, "weight"])
  } else if (wcgs[j,"typchd69"] == uniqueTypch69[3,1]) { # matches third label
    silentMIWeight <- append(silentMIWeight, wcgs[j, "weight"])
  } else if (wcgs[j,"typchd69"] == uniqueTypch69[4,1]) { # matches fourth label
    anginaWeight <- append(anginaWeight, wcgs[j, "weight"])
  } else {
    print("ERROR") # safeguard
  }
}

mean(noCHDWeight)
mean(miOrSDWeight) 
mean(silentMIWeight) 
mean(anginaWeight) 

```
B. Now do the same thing, but now don’t use a loop

``` {r echo = TRUE}
library(dplyr)

wcgs %>%
  group_by(typchd69) %>%
  summarise(meanWeight = mean(weight), sdWeight = sd(weight))

```
